Turbulence and Particle Acceleration in Giant Radio Halos: the Origin of Seed Electrons

### 2016-04-20 paper.tex
Added files via upload

\author[A. Pinzke, S. Peng Oh and C. Pfrommer] 
{Anders Pinzke$^{1,2}$\thanks{apinzke@fysik.su.se (AP); peng@physics.ucsb.edu (SPO); christoph.pfrommer@h-its.org (CP)}, S. Peng Oh$^{3}$ and Christoph Pfrommer$^{4}$\footnotemark[1]\\
$^{1}$The Oskar Klein Centre for Cosmoparticle Physics, Department
  of Physics, Stockholm University, AlbaNova University Center, SE - 106 91
  Stockholm, Sweden\\
$^{2}$Dark Cosmology Center, University of Copenhagen,
  Juliane Maries Vej 30, DK-2100 Copenhagen, Denmark\\
  $^{3}$University of California - Santa Barbara,
  Department of Physics, CA 93106-9530, USA\\
$^{4}$Heidelberg Institute for Theoretical Studies
  (HITS), Schloss-Wolfsbrunnenweg 35, 69118 Heidelberg, Germany}

@@

\begin{figure*}
  \includegraphics[width=1.0\columnwidth]{fCR.radius.coma.g72a.Rad14.2400p.z0.NL.xKR.eb23.eI067.DII.140.v6.eps}
  \caption{Spatial distribution of CRp energy density in the Coma
    cluster. The red dash-dotted line shows the required distribution
    of seed CRps that generate secondary electrons via p-p collisions
    required to reproduce Coma radio brightness observations after
    Fermi-II reacceleration \citep{brunetti12}. The blue solid line
    shows the distribution of fossil CRps found in cosmological
    simulations, which disagrees with the required profile. To better
    compare the two models in this figure, we normalize the required
    distribution of CRps by fixing the total CRp energy $E_\rmn{CR}$
    to 0.3 percent of the total thermal energy, consistent with
    observations \citep{2014ApJ...787...18A,2012ApJ...757..123A}.}
  \label{fig:Edens}
\end{figure*}

@@

\begin{figure*}
  \includegraphics[width=1.0\columnwidth]{turb_profile_ratio_tot.eps}
  \caption{The ratio of turbulent-to-thermal energy densities (solid
    lines) and cumulative energies (dotted lines) in our three
    models. The energy densities are parametrized as
    $\epsilon_\rmn{turb} \propto
    \epsilon_\rmn{th}^{(\alpha_{\rmn{tu}}+1)/2}/T^{1/4}$ and
    normalized such that the total turbulent energy in compressible
    modes $E_\rmn{turb}$ for each scenario makes up about 6\% of the
    total thermal energy $E_\rmn{th}$ inside the radio halo
    ($R_\rmn{RH}\approx0.6R_{200}$). The turbulent profiles explore
    the uncertainty in the cluster turbulence and are motivated by the
    cosmological simulation in
    \citep{2009ApJ...705.1129L,2010ApJ...725.1452S,2011A&A...529A..17V}.}
  \label{fig:Edens}
\end{figure*}

@@

\begin{table}
  \caption{Thermal quantities and timescales for different spatial
    regions in a Coma like cluster.}
\begin{tabular}{l c  c c}
\hline
\hline
& & spatial regions & \\
 & $0.1\,\RH^{(2)}$ & $0.3\,\RH^{(2)}$ & $\RH^{(2)}$   \\
\hline
thermal quantities$^{(1)}$ & & & \\
$\rho$ [$10^{-27}$ g cm$^{-3}$] & 3.0 & 1.6 & 0.15 \\
$T$ [$10^{8}$ K] & 1.4 & 1.0 & 0.58 \\
\hline
timescales$^{(1)}$ & & & \\
$\tau_\rmn{D}$(\Mprimary)$^{(3)}$ [Gyr] & 0.45 & 0.44 & 0.39 \\
$\tau_\rmn{D}$(\Mstream)$^{(3)}$  [Gyr] & 0.50 & 0.47 & 0.34 \\
$\tau_\rmn{D}$(\Mflatturb)$^{(3)}$  [Gyr] & 0.69 & 0.56 & 0.27 \\
$\tau_\rmn{IC/sync}(P=10^4\,m_ec)^{(4)}$ [Gyr] & 0.11 & 0.15 & 0.22 \\
$\tau_\rmn{pp}(P=100\,m_pc)^{(5)}$ [Gyr] & 2.4 & 4.5 & 47 \\
$\tau_\rmn{coul}(P=\,m_ec)^{(6)}$  [Gyr] & 0.0092 & 0.017 & 0.17 \\
\hline
\end{tabular}
\begin{quote}
 Notes: \\ 
 (1) Median quantities from our simulated post-merging cluster g72a.\\
 (2) Radius of the giant radio halo in Coma where $\RH\approx0.6\,R_{200}$.\\
 (3) Fermi-II reacceleration for both electrons and protons at all energies.\\
 (4) Inverse Compton and Synchrotron cooling for electrons.\\
 (5) Catastrophic losses for protons.\\
 (6) Coulomb cooling for electrons (protons factor $m_e/m_p$ smaller).\\

@@

\section{Parameter space exploration with spherical models}
...
\begin{figure*}
\begin{minipage}{1\columnwidth}
   \begin{center}\Large{Radio profiles}\\
     \includegraphics[width=\columnwidth]{prof.comp.KrTTDth.Xtu.eps}
   \end{center}
\end{minipage}
\begin{minipage}{1\columnwidth}
   \begin{center}\Large{Radio spectra}\\
     \includegraphics[width=\columnwidth]{spec.comp.KrTTDth.Xtu.eps}
   \end{center}
\end{minipage}
\\
\begin{minipage}{1\columnwidth}
  \begin{center}%\Large{\Mflatturb:}\\ 
    \includegraphics[width=\columnwidth]{prof.comp.KrTTDth.aCR.eps}
  \end{center}
\end{minipage}
\begin{minipage}{1\columnwidth}
   \begin{center}%\Large{\it Brunetti et al. (2012)}:\\
     \includegraphics[width=\columnwidth]{spec.comp.KrTTDth.aCR.eps}
   \end{center}
\end{minipage}
\\
\begin{minipage}{1\columnwidth}
  \begin{center}%\Large{\Mflatturb:}\\ 
    \includegraphics[width=\columnwidth]{prof.comp.KrTTDth.aI0.eps}
  \end{center}
\end{minipage}
\begin{minipage}{1\columnwidth}
   \begin{center}%\Large{\it Brunetti et al. (2012)}:\\
     \includegraphics[width=\columnwidth]{spec.comp.KrTTDth.aI0.eps}
   \end{center}
\end{minipage}
\caption{...}
  \label{fig:param_comp}
\end{figure*}

@@

\section{Results of cosmological simulations}
\label{sec:results}
In this section we show that our three models individually reproduce
the observed Coma radio halo without violating gamma-ray bounds. The
level of emission is mainly driven by how efficient the CRs are
reaccelerated and for how long. The diffusion constant,
$D_\rmn{pp}\propto I_0/\rho c_s \propto
X_\rmn{tu}^2k_0\epsilon_\rmn{th}^{\alpha_\rmn{tu}-1}\sqrt{T}$, where
we adopt the same level of compressible turbulence, $X_\rmn{tu}=0.06$,
for all models. The reaccelerated CRs are super-exponentially
dependent on this parameter as long as $\tau_D > \tau_\rmn{cl}$, where
$\tau_\rmn{cl}$ represent the duration of reacceleration of the merger
which we assume to be 650~Myrs long. Note that fixing $D_\rmn{pp}$
allows us to relate $\tau_\rmn{cl} \propto X_\rmn{tu}^2 \propto k_0$
that ultimately sets the normalization for $D_\rmn{pp}$. We also
exploit the large uncertainty in the turbulent profile and adjust
$\alpha_\rmn{tu}$ for each model so that the CRs have the correct
spatial profile to match the radio emission. The shape of the spectrum
for each of model is more complicated and is determined both from
reacceleration and as well as the gains and losses during the
formation of the cluster. It depends only weakly on the specifics of
the CR injection at the shocks (where the Mach number distribution of
the shocks is the most important quantity). The normalization of the
spectrum is mainly determined from the maximum acceleration efficiency
(together with the uncertain parameters $X_\rmn{tu}$, $\tau_\rmn{cl}$,
and $k_0$). Given the freedom in the turbulens, it might not be
surprising that we match the radio profile in Coma. Although, matching
the spectrum without violating gamma-ray bounds for each of our three
models is more surprising and is not a consequence of parameter
fitting. We believe that it is ultimately a consequence of the adopted
physics and hint that reacceleration is a viable explanation for giant
radio halos. Below we discuss in more detail the different emission
components of Coma.

>>>>

\author[A. Pinzke, S. Peng Oh and C. Pfrommer] 
{Anders Pinzke$^{1,2}$\thanks{apinzke@fysik.su.se (AP); peng@physics.ucsb.edu (SPO); christoph.pfrommer@h-its.org (CP)}, S. Peng Oh$^{3}$ and Christoph Pfrommer$^{4}$\footnotemark[1]\\
$^{1}$The Oskar Klein Centre for Cosmoparticle Physics, Stockholm University, AlbaNova University Center, SE - 106 91
  Stockholm, Sweden\\
$^{2}$Dark Cosmology Center, University of Copenhagen,
  Juliane Maries Vej 30, DK-2100 Copenhagen, Denmark\\
  $^{3}$University of California - Santa Barbara,
  Department of Physics, CA 93106-9530, USA\\
$^{4}$Heidelberg Institute for Theoretical Studies
  (HITS), Schloss-Wolfsbrunnenweg 35, 69118 Heidelberg, Germany}

@@

\begin{figure}
  \includegraphics[width=1.0\columnwidth]{fCR.radius.coma.g72a.Rad14.2400p.z0.NL.xKR.eb23.eI067.DII.140.v6.eps}
  \caption{Spatial distribution of CRp energy density in the Coma
    cluster. The red dash-dotted line shows the required distribution
    of seed CRps that generate secondary electrons via p-p collisions
    required to reproduce Coma radio brightness observations after
    Fermi-II reacceleration \citep{brunetti12}. The blue solid line
    shows the distribution of fossil CRps found in cosmological
    simulations, which disagrees with the required profile. To better
    compare the two models in this figure, we normalize the required
    distribution of CRps by fixing the total CRp energy $E_\rmn{CR}$
    to 0.3 percent of the total thermal energy, consistent with
    observations \citep{2014ApJ...787...18A,2012ApJ...757..123A}.}
  \label{fig:Edens}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=1.0\columnwidth]{turb_profile_ratio_tot.eps}
  \caption{The ratio of turbulent-to-thermal energy densities (solid
    lines) and cumulative energies (dotted lines) in our three
    models. The energy densities are parametrized as
    $\epsilon_\rmn{turb} \propto
    \epsilon_\rmn{th}^{(\alpha_{\rmn{tu}}+1)/2}/T^{1/4}$ and
    normalized such that the total turbulent energy in compressible
    modes $E_\rmn{turb}$ for each scenario makes up about 6\% of the
    total thermal energy $E_\rmn{th}$ inside the radio halo
    ($R_\rmn{RH}\approx0.6R_{200}$). The turbulent profiles explore
    the uncertainty in the cluster turbulence and are motivated by the
    cosmological simulation in
    \citep{2009ApJ...705.1129L,2010ApJ...725.1452S,2011A&A...529A..17V}.}
  \label{fig:Edens}
\end{figure}

@@

\begin{table}
  \caption{Thermal quantities and timescales for different spatial
    regions in a Coma like cluster.}
\begin{tabular}{l c  c c}
\hline
\hline
& & spatial regions & \\
 & $0.1\,\RH^{(2)}$ & $0.3\,\RH^{(2)}$ & $\RH^{(2)}$   \\
\hline
thermal quantities$^{(1)}$ & & & \\
$\rho$ [$10^{-27}$ g cm$^{-3}$] & 3.0 & 1.6 & 0.15 \\
$T$ [$10^{8}$ K] & 1.4 & 1.0 & 0.58 \\
\hline
timescales$^{(1)}$ & & & \\
$\tau_\rmn{D}$(\Mprimary)$^{(3)}$ [Gyr] & 0.45 & 0.44 & 0.39 \\
$\tau_\rmn{D}$(\Mstream)$^{(3)}$  [Gyr] & 0.50 & 0.47 & 0.34 \\
$\tau_\rmn{D}$(\Mflatturb)$^{(3)}$  [Gyr] & 0.69 & 0.56 & 0.27 \\
$\tau_\rmn{IC/sync}(P=10^4\,m_ec)^{(4)}$ [Gyr] & 0.11 & 0.15 & 0.22 \\
$\tau_\rmn{pp}(P=100\,m_pc)^{(5)}$ [Gyr] & 2.4 & 4.5 & 47 \\
$\tau_\rmn{coul}(P=\,m_ec)^{(6)}$  [Gyr] & 0.0092 & 0.017 & 0.17 \\
\hline
\end{tabular}
\begin{quote}
 Notes: \\ 
 (1) Median quantities from our simulated post-merging cluster g72a.\\
 (2) Radius of the giant radio halo in Coma where $\RH\approx0.6\,R_{200}$.\\
 (3) Fermi-II reacceleration for both electrons and protons at all energies.\\
 (4) Inverse Compton and synchrotron cooling for electrons.\\
 (5) Catastrophic losses for protons.\\
 (6) Coulomb cooling for electrons (protons factor $m_e/m_p$ smaller).\\

@@

\section{Parameter space exploration with spherical models}
In this section we explore the robustness of our models to crucial
parameters. We do not trace single Lagrangian particles in the
cluster, instead we use a more simple approach where we solve the
Fokker-Plack equation in static spherical shells for injection,
reacceleration, and losses of the CRs. Once the CRs have been
reaccelerated for $\tau_\rmn{cl} = 650\,\rmn{Myr}$, we calculate the
radio emission and compare the profile and spectra for different
parametrizations.

We adopt both the density \cite{1992A&A...259L..31B} and temperature
\cite{2009ApJ...696.1886B,2001A&A...365L..67A} profiles derived from
X-ray observations of the Coma cluster. They are given by:
\begin{eqnarray}
n_e = \frac{3.4\times10^{-3}}{\left[1+\left(R/294\,\rmn{kpc}\right)^2\right]^{1.125}}\nonumber\\
T k_B = 8.25\,\rmn{keV} \left[1+\left(R/0.5R_{200}\right)^2\right]^{-0.32}\,.
\end{eqnarray}
The virial radius of Coma is given by $R_{200} = 2300\,\rmn{kpc}$.  The
bulk of the CRps are injected by relatively low Mach number shocks and
parametrized by $f_\rmn{CRp,inj}(p) = C_\rmn{CRp,inj}\,p^(-2.5)$. The
CRps trace the thermal gas with $C_\rmn{CRp,inj} \propto
\epsilon_\rmn{th}^{\alpha_\rmn{CR,spat}}$ where the normalization is fixed
by assuming that the injected CR energy in the last 650~Myrs make up
0.03\% of the thermal energy inside the virial radius, i.e.
$\int\epsilon_\rmn{CRp,inj} d\rmn{V}/\int\epsilon_\rmn{th} d\rmn{V} =
0.0003$. The spectrum of the initial CRp distibtion is determined by
is given by the steady state between injection and cooling,
\begin{equation}
 f_{CRp,0} \approx \frac{\int_p^\infty f_\rmn{CRp,inj}(p') 
\rmn{d}p'}{\left|{{dp}\over{dt}}\right|_{\rm C}+\frac{p}{\tau_{\rm had}}}\,,
\end{equation}
where the normalization is determined from $\int\epsilon_\rmn{CRp,inj}
d\rmn{V}/\int\epsilon_\rmn{th} d\rmn{V} = 0.003$. Similarly, the
initial CRe distribution is given by the steady state between cooling
(Coulomb, invers Compton, and synchroton) and injection from
$f_{CRp,0}$. We also include the contineous injection of secondary
CRes during $\tau_\rmn{cl}$.

The compressible turbulence responsible for reacceleating both the CRps and
CRes is parametrized by $I_0 \propto X_\rmn{tu} (n_e
T)^\alpha_\rmn{tu}$. The diffusion constant $D_\rmn{pp}$ is 
calculated for each radial bin using
Eqns.~\ref{eq:Gamma_e}-\ref{eq:dpp}.

In Fig.~\ref{fig:param_comp} we show how assumptions about turbulence
and CRs impact the radio emission. We find that the level of
turbulence has a large impact on both radio spectra and profile, while
the parameter $\alpha_\rmn{tu}$ that determines the spatial profile of
turbulence mainly influence the radio profiles. Interestingly,
assumptions about CRs have less impact on radio emission. This can be
understood by again looking at the scaling of parameters to the
reacceleration timescale, where $\tau_\rmn{D} \propto
X_\rmn{tu}^2$. The CRps approximately gain a factor $\Delta
C_\rmn{reacc.} \sim
\exp{\frac{(2+\alpha)(1-\alpha)}{4}\frac{\tau_\rmn{cl}}{\tau_\rmn{D}}}$
during reacceleration if injection and cooling are neglected. The CRes
gain a similar factor from reacceleration, however in the region
around $0.1\,R_{200}$ the Coulomb cooling timescale
$\tau_\rmn{coul}\approx0.01\,\rmn{Myr}$ is much shorter than
$\tau_\rmn{D}\approx0.5\,\rmn{Myr}$, hence $\Delta C_\rmn{reacc.}$
will be greatly reduced for the electrons. The spatial region is
important when we derive the radio spectrum, since in order to compare
to observations we integrate the CR profile out to $\sim
0.2\,R_{200}$. For a CRp spectral index $\alpha_\rmn{CR}\sim 2.5$,
$\Delta C_\rmn{reacc.} \sim
\exp{1.7\frac{\tau_\rmn{cl}}{\tau_\rmn{D}}} \sim 10$ for
$X_\rmn{tu}=0.2$ and $\alpha_\rmn{tu}=0.8$. This is confirmed by the
top panels in Fig.~\ref{fig:param_comp}. Increasing from
$X_\rmn{tu}=0.2$ to $X_\rmn{tu}=0.3$, gives us a $\Delta
C_\rmn{reacc.}  \sim 100$, which is roughly confirmed by same
figures. The difference is due to the shorter reacceleration timescale
that increase the contribution from reaccelerated CRes with a factor
few. \AP{Too complicated, perhaps cut or simplify?}

\begin{figure*}
\begin{minipage}{1\columnwidth}
   \begin{center}\Large{Radio profiles}\\
     \includegraphics[width=\columnwidth]{prof.comp.KrTTDth.Xtu.eps}
   \end{center}
\end{minipage}
\begin{minipage}{1\columnwidth}
   \begin{center}\Large{Radio spectra}\\
     \includegraphics[width=\columnwidth]{spec.comp.KrTTDth.Xtu.eps}
   \end{center}
\end{minipage}
\\
\begin{minipage}{1\columnwidth}
  \begin{center}%\Large{\Mflatturb:}\\ 
    \includegraphics[width=\columnwidth]{prof.comp.KrTTDth.aI0.eps}
  \end{center}
\end{minipage}
\begin{minipage}{1\columnwidth}
   \begin{center}%\Large{\it Brunetti et al. (2012)}:\\
     \includegraphics[width=\columnwidth]{spec.comp.KrTTDth.aI0.eps}
   \end{center}
\end{minipage}
\\
\begin{minipage}{1\columnwidth}
  \begin{center}%\Large{\Mflatturb:}\\ 
    \includegraphics[width=\columnwidth]{prof.comp.KrTTDth.aCR.eps}
  \end{center}
\end{minipage}
\begin{minipage}{1\columnwidth}
   \begin{center}%\Large{\it Brunetti et al. (2012)}:\\
     \includegraphics[width=\columnwidth]{spec.comp.KrTTDth.aCR.eps}
   \end{center}
\end{minipage}
\caption{Sensitivity of radio emission in Coma cluster to critical
  parameters. {\it Left panels} show the radio surface brightness
  profiles. We compare profiles at 352~MHz \citep[blue lines and
    crosses,][]{brown11} to predicted emission from Fermi-II
  reaccelerated CR electrons populations (black lines). {\it Right
    panels} show radio synchrotron spectra. The green crosses are
  compiled from observations \citet{2010PhDT.......259P}, while the
  black lines show predicted emission from reaccelerated CR
  electrons. The upper panels show the sensitivity to the level of
  turbulence ($X_\rmn{tu}$), middle panels show the impact of
  different turbulent profiles ($\alpha_\rmn{tu}$), and lower panels
  show the dependence on spatial distributions of initial and injected
  CRs ($\alpha_\rmn{CR,spat}$). We adopt the following fiducial values
  for parameters that are not varied in each panel: $X_\rmn{tu}=0.2$,
  $\alpha_\rmn{tu}=0.8$, and $\alpha_\rmn{CR,spat}$. Note that the
  radio emission is much more sensitive to assumptions about turbulent
  parameters than CRs.}
  \label{fig:param_comp}
\end{figure*}

@@

\section{Results of cosmological simulations}
\label{sec:results}
In this section we show that our three models individually reproduce
the observed Coma radio halo without violating gamma-ray bounds. The
level of emission is mainly driven by how efficient the CRs are
reaccelerated and for how long. The diffusion constant,
$D_\rmn{pp}\propto I_0/\rho c_s \propto
X_\rmn{tu}^2k_0\epsilon_\rmn{th}^{\alpha_\rmn{tu}-1}\sqrt{T}$, where
we adopt the same level of compressible turbulence, $X_\rmn{tu}=0.06$,
for all models. The reaccelerated CRs are exponentially dependent on
this parameter as long as $\tau_D > \tau_\rmn{cl}$, where
$\tau_\rmn{cl}$ represent the duration of reacceleration of the merger
which we assume to be 650~Myrs long. Note that fixing $D_\rmn{pp}$
allows us to relate $\tau_\rmn{cl} \propto X_\rmn{tu}^2 \propto k_0$
that ultimately sets the normalization for $D_\rmn{pp}$. We also
exploit the large uncertainty in the turbulent profile and adjust
$\alpha_\rmn{tu}$ for each model so that the CRs have the correct
spatial profile to match the radio emission. The shape of the spectrum
for each of model is more complicated and is determined both from
reacceleration and as well as the gains and losses during the
formation of the cluster. It depends only weakly on the specifics of
the CR injection at the shocks (where the Mach number distribution of
the shocks is the most important quantity). The normalization of the
spectrum is mainly determined from the maximum acceleration efficiency
(together with the uncertain parameters $X_\rmn{tu}$, $\tau_\rmn{cl}$,
and $k_0$). Given the freedom in the turbulens, it might not be
surprising that we match the radio profile in Coma. Although, matching
the spectrum without violating gamma-ray bounds for each of our three
models is more surprising and is not a consequence of parameter
fitting. We believe that it is ultimately a consequence of the adopted
physics and hint that reacceleration is a viable explanation for giant
radio halos. Below we discuss in more detail the different emission
components of Coma.
